Staged Abstract Interpreters

### 2019-08-13 paper/50_stageabs.tex
revision

Our staging schema works well in theory, but in practice would suffer from code
explosion, and possibly runtime GC overhead (at the next stage) when analyzing
(specializing) large programs. In this section, we present several optimizations
that largely mitigate these issues. Still, implementing these optimizations do
not need to change the generic interpreter.

@@

\todo{TODO}

@@

\todo{TODO}

As one of the merits of our approach, the staging does not compromise any
soundness, because we only change the underlying monads and data structures to
the staged version, not any part of the abstract semantics. Based on the
assumption that MSP system and staged data structure preserve the equivalence
during staging, we are confident that the staged abstract interpreter does the
same analysis as the unstaged one.

>>>>

Our staging schema works by unfolding the interpreter over the input AST. In
practice, it would suffer from code explosion when analyzing (specializing) large
programs, which increases the compile time.
If we generate next stage programs running on JVM, such large generated program
would also cause runtime GC overhead. In this section, we present several
optimizations that largely mitigate these issues. For all that, implementing
these optimizations do not need to change the generic interpreter.

@@

\todo{1. construct generic monadic interpreter that abstract over binding-time,
  primtive operations, return type.
  2. implement unstaged abstract interpreter with monads, ie. using approriate
  monad stack
  3. if we want to generate code, we just need to change the monad stack.
  the data wrapped are staged.
  we show readerT monad, stateT monad, fused monad with nondeterminism. }

@@

As the central concern of static analysis, the soundness is important for
every user who would like to use it. Here we briefly summarize that how our
approach preserve the corretness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here are based on empirical evidence, a rigorous proof of soundness
perservation of the staged abstract interpreters is still an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistent, i.e., the generated program does not change the observational
    behaviours of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal form that preserves the
    execution order within a stage \todo{cite WF16}. However, in general, it is
    possible to subvert these guarantees by using unsafe features, such as
    casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularlly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structure we used here are simply
    a black-box that wraps the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

### 2019-08-13 paper/80_related.tex
revision

\paragraph{Optimizing Static Analysis Through Specialization}
The underlying idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretation overhead on
traversing the syntax tree by specialization. Specifically, the residual program
of AC can be either textual or closures. However, as we studied in Section
\ref{cs_ac}, with the perspective from generative programming and monadic
interpreters, we make it more general, extensible and easy.
\citet{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure
generation to optimize small-step abstract interpreter in state-transition
style. The analyzed program is firstly compiled to an IR called "abstract
bytecode", which are actually higher-order functions and will be executed later
on an abstract abstract machine for that IR. \citet{damian1999partial} provides a
formal treatment to abstract compilation and Shiver's CFA, as well as proofs to
establish the correctness of the certified specialized analyzer.
\citet{amtoft1999partial} applied partial evaluation for constraint-based
control flow analysis. Splitting an analysis into multiple stages is also
studied for analysis other than for control-flow, though the formulation may
very different. For example, \citet{DBLP:conf/cgo/HardekopfL11} apply staging to
flow-sensitive pointer analysis. The first stage is to analyze the program code
to obtain a sparse representation, and then the second stage conducts the
flow-sensitive analysis based on the first one.
The abstract domains can be also specialized w.r.t. the static program
structure, e.g., decomposing polyhedras \cite{DBLP:conf/popl/SinghPV17,
Singh:2017:PCD:3177123.3158143} to smaller ones depending on the variables
involved such that the abstract transformers can compute using less time and
space.

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As to build semantic artifacts, the Abstracting
Abstract Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
approach shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The AAM approach is closely related to control-flow analysis
for higher-order languages \cite{Midtgaard:2012:CAF:2187671.2187672}.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting definitional
interpreters approach from \citet{DBLP:journals/pacmpl/DaraisLNH17}.
Similar to definitional abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis, \todo{but intentional properties
  are different, compare with summary}

>>>>

\paragraph{Optimizing Static Analysis Through Specialization}
The underlying idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretation overhead on
traversing the syntax tree by specialization. Specifically, the residual program
of AC can be either textual or closures. However, as we studied in Section
\ref{cs_ac}, with the perspective from generative programming and monadic
interpreters, we make it more general, extensible and easy.
\citet{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure
generation to optimize small-step abstract interpreter in state-transition
style. The analyzed program is firstly compiled to an IR called "abstract
bytecode", which are actually higher-order functions and will be executed later
on an abstract abstract machine for that IR. \citet{damian1999partial} provides a
formal treatment to abstract compilation and Shiver's CFA, as well as proofs to
establish the correctness of the certified specialized analyzer.
\citet{amtoft1999partial} applied partial evaluation for constraint-based
control flow analysis.

Splitting an analysis into multiple stages is also studied for analysis other
than abstract-interpretation-based control-flow analysis, though the formulation may
very different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the program code
to obtain a sparse representation, and then the second stage conducts the
flow-sensitive analysis by refining the first one. Another large class of static
analysis are implemented by generating constraints firstly and then let a solver
discharge these proof obligations or find the fixed-point. Examples include SAT-
and SMT-based program analysis\todo{cite}, Datalog-based points-to
analysis\todo{cite}, and \todo{}. Similar to the idea in this paper, such
analyses can be splitted to staged.
One of the future work is to explore that whether these analyses can be
implemented and optimized from metaprogramming. Souffle, one of the
state-of-the-art Datalog engine that is commonly used in points-to analysis,
uses the idea of Futamura projection for efficent implemetation. But it only
specializes Datalog rules to the running engine, instead of a full pipeline from
program AST to generating constraints and solving constraints.

Recent work shows that the numerical abstract domains can be also specialized
w.r.t. the static program structure. For example, decomposing polyhedras
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to smaller ones
depending on the variables involved in a statement can significantly improve the
running time and space of abstract transformers.

\paragraph{Construction of Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As to build semantic artifacts, the Abstracting
Abstract Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
approach shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The AAM approach is closely related to control-flow analysis
for higher-order languages \cite{Midtgaard:2012:CAF:2187671.2187672}.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting definitional
interpreters approach from \citet{DBLP:journals/pacmpl/DaraisLNH17}.
Similar to definitional abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.

\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis, \todo{but intentional properties
  are different, compare with summary}
